Formalizing the Kolmogorov extension Theorem in Lean

### 2023-09-11 Manuscript/proof_outline.tex
update manuscript

{\color{blue}
  \noindent
  TODO
  \begin{itemize}
  \item Make sure that the line of argument for code and mauscript are in line.
  \end{itemize}
}

@@

\begin{lemma}\label{l2}
  Let $\mathcal H$ be a semi-ring and $A_1,...,A_m \in \mathcal
  H$. Then, there are $\mathcak K_1,...,\mathcal K_m \subseteq_f
  \mathcal H$ disjoint such that $\bigcup_{n=1}^m \mathcal K_n$
  contains disjoint sets and $\bigcup_{m=1}^n A_m = \biguplus_{m=1}^n
  \biguplus_{K \in \mathcal K_n} K$.  
\end{lemma}

@@

\noindent
In the proof of Kolmogorov's extension theorem, we use a closely
related definition of \leanline{Measure.ofAddContent}, with variables
\leanline{[mα : MeasurableSpace α] (hC : SetSemiring C) (hC_gen :
  MeasurableSpace.generateFrom C = mα)}, i.e.\ the semi-ring
\leanline{C} generates the $\sigma$-algebra, and which uses an
\leanline{m : AddContent} as well as \leanline{m_sigma_subadd} as
above.

@@

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem kolContent_sigma_additive_of_innerRegular
  (hP : IsProjectiveMeasureFamily P)
  (hP_inner : ∀ J,
    (P J).InnerRegular (fun s => IsCompact s ∧ IsClosed s)
    MeasurableSet)
  ⦃f : ℕ → Set (∀ i, α i)⦄ (hf : ∀ i, f i ∈ cylinders α)
  (hf_Union : (⋃ i, f i) ∈ cylinders α)
  (h_disj : Pairwise (Disjoint on f)) :
  kolContent hP (⋃ i, f i) = ∑' i, kolContent hP (f i)
\end{minted}

@@

We apply this result by using that 1.\ is satisfied for complete and
separable extended pseudo-metric spaces:

@@

\subsection{Proof of Kolmogorov's extension theorem}
With the above results, the strategy of the proof of Theorem~\ref{T1}
is as follows: We have to show that under the given conditions,
condition~1 of Lemma~\ref{l:tight} holds. By the equivalent~2., we
find that $P$ is inner regular with respect to closed compact sets. In
particular, the conditions of Lemma~\ref{l:stetigcompact} are
satisfied, showing that $P$ is $\sigma$-additive. Then, we proceed as
described right before Proposition~\ref{P:stetigmass}. So, it remains
to show condition~1 of Lemma~\ref{l:tight} for complete and separable
extended pseudo-metric spaces. Note that -- as a special case of
\leanline{isCompact_iff_totallyBounded_isComplete} from {\tt
  mathlib}, any subset of such a space is compact iff it is complete
and totally bounded. Since closed subsets of a complete space are
complete, the closure of a totally bounded set is (still totally
bounded, hence) compact.

Now we can put everything together: According to
Lemma~\ref{L:relcoPol}, condition~1 of Lemma~\ref{l:tight} is
satisfied for the set function $P$ on the ring $\mathcal A$ from
Theorem~\ref{T1}. The same lemma shows that the conditions for
Lemma~\ref{l:stetigcompact} are satisfied, so $P$ is $\sigma$-additive
on $\mathcal A$. As a consequence, we can use the Carathéodory
extension from Theorem~\ref{T:masseind}. This shows all other
assertions. The formalization of this proof resembles these
arguments. We leave out all instances in the reformulation of the
result and its proof (see below Theorem~\ref{T1} for a full
formulation):

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  noncomputable def projectiveLimitWithWeakestHypotheses
    [∀ i, PseudoEMetricSpace (α i)]
    [∀ i, BorelSpace (α i)]
    [∀ i, TopologicalSpace.SecondCountableTopology (α i)]
    [∀ i, CompleteSpace (α i)] [Nonempty (∀ i, α i)]
    (P : ∀ J : Finset ι, Measure (∀ j : J, α j))
    [∀ i, IsFiniteMeasure (P i)]
    (hP : IsProjectiveMeasureFamily P) :
    Measure (∀ i, α i) :=
      Measure.ofAddContent setSemiringCylinders
      generateFrom_cylinders (kolContent hP)
      (kolContent_countably_subadditive_of_innerRegular hP
        fun J => innerRegular_isCompact_isClosed_
        measurableSet_of_complete_countable (P J))    
\end{minted}

>>>>

@@

\begin{lemma}\label{l2}
  Let $\mathcal H$ be a semi-ring and $A_1,...,A_m \in \mathcal
  H$. Then, there are $\mathcal K_1,...,\mathcal K_m \subseteq_f
  \mathcal H$ disjoint such that $\bigcup_{n=1}^m \mathcal K_n$
  contains disjoint sets and $\bigcup_{m=1}^n A_m = \biguplus_{m=1}^n
  \biguplus_{K \in \mathcal K_n} K$.  
\end{lemma}

@@

\noindent
In the proof of Kolmogorov's extension theorem, we use a closely
related definition:

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
noncomputable def Measure.ofAddContent
  [mα : MeasurableSpace α] (hC : SetSemiring C)
  (hC_gen : MeasurableSpace.generateFrom C = mα)
  (m : AddContent C)
  (m_sigma_subadd : ...) :
  Measure α
\end{minted}
with \leanline{m_sigma_subadd} as above.

@@

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem kolContent_sigma_additive_of_innerRegular
  (hP : IsProjectiveMeasureFamily P)
  (hP_inner : ∀ J,
    (P J).InnerRegular (fun s => IsCompact s ∧ IsClosed s)
    MeasurableSet)
  ⦃f : ℕ → Set (∀ i, α i)⦄ (hf : ∀ i, f i ∈ cylinders α)
  (hf_Union : (⋃ i, f i) ∈ cylinders α)
  (h_disj : Pairwise (Disjoint on f)) :
  kolContent hP (⋃ i, f i) = ∑' i, kolContent hP (f i)

  theorem kolContent_countably_subadditive_of_innerRegular
    (hP : IsProjectiveMeasureFamily P)
    (hP_inner : ∀ J,
      (P J).InnerRegular (fun s => IsCompact s ∧ IsClosed s)
      MeasurableSet)
    ⦃f : ℕ → Set (∀ i, α i)⦄
    (hf : ∀ i, f i ∈ cylinders α)
    (hf_Union : (⋃ i, f i) ∈ cylinders α) :
    kolContent hP (⋃ i, f i) ≤ ∑' i, kolContent hP (f i)
\end{minted}

@@

We apply this result by using that 1.\ is satisfied for complete and
separable extended pseudo-metric spaces. Here, for the generalization
to extended pseudo-metric spaces, we use that any subset of such a
space is compact iff it is complete and totally bounded (see
\leanline{isCompact_iff_totallyBounded_isComplete} in {\tt
  mathlib}). Since closed subsets of a complete space are complete,
the closure of a totally bounded set is (still totally bounded, hence)
compact. (This fact is used in the proof below.)

@@

\subsection{Proof of Kolmogorov's extension theorem}
We now describe the proof of Kolmogorov's extension theorem as well as
its formalization: For the proof of Theorem~\ref{T1}, we 
\begin{enumerate}
  \item apply Theorem~\ref{T:masseind} for the ring (hence semi-ring)
    $\mathcal A$ and the set-function $P$ as given in
    Theorem~\ref{T1};
  \item show $\sigma$-additivity (as assumed in
    Theorem~\ref{T:masseind}) of $P$ by using
    Lemma~\ref{l:stetigcompact}. For the latter, we need to show that
    $P$ is inner regular with respect to a compact system. Here, note
    that $\{\pi_J^{-1}C : C \in \prod_{j\in J} \alpha_j \text{ compact
      and closed}\}$ is a compact system and $P(\pi_J^{-1}C) =
    P_J(C)$, so we need to show that $P_J$ is inner regular with
    respect to compact and closed sets, $J\subseteq_f \iota$;
  \item use Lemma~\ref{L:relcoPol} in combination with
    Lemma~\ref{l:tight} (1.$\Rightarrow$2.) and the properties of
    completeness, separability of the underlying extended
    pseudo-metric spaces in order to see that every $P_J$ has the
    desired property of being inner regular wrt compact and closed
    sets. This concludes the proof of Theorem~\ref{T1}. 
\end{enumerate}
The formalization of this proof resembles these arguments. We leave
out all instances in the reformulation of the result and its proof
(see below Theorem~\ref{T1} for a full formulation):

\begin{minted}[mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
noncomputable def
  projectiveLimitWithWeakestHypotheses
  (P : ∀ J : Finset ι, Measure (∀ j : J, α j))
  (hP : IsProjectiveMeasureFamily P) :
  Measure (∀ i, α i) :=
    Measure.ofAddContent setSemiringCylinders
    generateFrom_cylinders (kolContent hP)
    (kolContent_countably_subadditive_of_innerRegular hP
      fun J => innerRegular_isCompact_isClosed_
      measurableSet_of_complete_countable (P J))    
\end{minted}
Let us look closer at the formalized proof:
\leanline{kolContent_countably_subadditive_of_innerRegular} has the
same hypothesis as connected to
\leanline{kolContent_sigma_additive_of_innerRegular}, except
\leanline{h_disj}. So, the content of the last bracket shows
$\sigma$-sub-additivity of the content $P$ from
Theorem~\ref{T1}. Then, \leanline{generateFrom_cylinders (kolContent
  hP)} gives the \leanline{MeasurableSpace} on which we define the
\leanline{addContent}.
